Reproducible Research assignment - Activity Monitoring
=====================================================
***
### Part 1 - Mean and Median of the Total number of steps taken per day.

***

#### Step 1 - Read the data. 


```{r}
act_data <- read.csv("activity.csv", header = T)
```

#### Step 2 - Group the data and summarize

Group the data by date and calculate sum of steps for each day using group_by and summarize functions of dplyr package

```{r}
library(dplyr)
daily <- act_data %>% 
          group_by(date) %>% 
           summarise(total_steps = sum(steps))

per_day <- daily$total_steps

```

#### Step 3 - Histogram of Total number of steps taken

```{r}
hist(per_day, 
     main = "Hisogram of Total Steps taken each Day",
     xlab = "Number of steps",
     col = "blue")

```

#### Step 4 - Calculate and report mean and median of steps

``` {r}
mean_steps <- mean(daily$total_steps, na.rm = T)
median_steps <- median(daily$total_steps, na.rm = T)
```
The mean of the the total steps taken is `r mean_steps`  
The median of the total steps is is `r median_steps`
***

### Part 2 - Average daily activity Pattern. 
            
***
#### Step 1 - Average daily activity Pattern 

This is to see the pattern of activity in different 5 minute intervals of time.  

First we group the data by interval and then calculate mean of each group using summarise function. 

```{r}

act_pat <- act_data %>% 
            group_by(interval) %>% 
            summarise(avg_steps = mean(steps, na.rm = T))

```

#### Step 2 - Time series plot of 5- minute interval

Here's a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}

plot(act_pat, type = "l", 
     main = "Average number of steps taken in each interval across all days",
     xlab = "5 minute Interval", 
     ylab = "Average number of steps", 
     col = "blue")

```

#### Step 2 - Interval with maximum mumber of steps

Calcualte the interval with maximum number of steps taken. 

```{r}

max_int <- act_pat[which.max(act_pat$avg_steps),1]
                   
```
Interval with maximum number of steps is `r max_int `.
***

### Part 3 - Imputing missing values
***
#### Step 1 - Calculate the number of missing values

```{r}

missing_values <- sum(is.na(act_data$steps))
```

The number of missing values is `r missing_values `

#### Step 2 - Remove the missing values, calculate daily activity and set the value of activity data to new object. 

```{r}
activity_rm <- act_data[complete.cases(act_data$steps),]
```

Calculate daily activity. 

```{r}

daily_act <- tapply(activity_rm$steps, activity_rm$interval, mean)

```
Set the value of original activity data to an object called new_act

``` {r}
new_act <- act_data
```

#### Step 3 - Impute the missing values in the new object. 

```{r}

new_act[which(is.na(new_act$steps)),1]<-
      daily_act[as.character(new_act[which(is.na(new_act$steps)),3])]

```

#### Step 4 - Calculate new daily activity with no missing values

```{r}
perday_new <- new_act %>% 
              group_by(date) %>% 
              summarise(total_steps = sum(steps)) 
```

#### Step 5 - Plot the data with and without missing values. 

```{r}

par(mfrow = c(1,2))

hist(per_day, 
     main = "Data with Missing values", 
     xlab = "Steps")

hist(perday_new$total_steps, 
     main = "Data without missing values", 
     xlab = "Steps")

```

              

#### Step 6 - Calculate mean and median of using imputed data

```{r}
mean_perdayNew <- mean(perday_new$total_steps)
median_perdayNew <- median(perday_new$total_steps)
```

Mean and Median of the new data is `r mean_perdayNew ` and `r median_perdayNew ` respectively.  

### Part 4 - Differences in activity patterns between weekdays and weekends

***

#### Step 1-  Setting the day of the week to each observation

Add a variable called day that converts the date to the corresponding day of the week

```{r}
new_act$day <- weekdays(as.Date(new_act$date))
```

 Add another factor variable called fday, which has two factors Weekend and Weekday
 
```{r} 
new_act$fday <- as.factor(c("Weekend", "Weekday"))
```


Set the values Weekend to Sunday and Saturday and Weekday to all others

```{r}
new_act[new_act$day == "Sunday" | new_act$day == "Saturday",5] <- factor("Weekend")
new_act[!(new_act$day == "Sunday" | new_act$day == "Saturday"),5] <- factor("Weekday")
```

#### Step 2 - Seperate the dataframe on Weekdays and Weekend

```{r}
new_act_we <- new_act[new_act$fday == "Weekend",]
new_act_wd <- new_act[new_act$fday == "Weekday",]
```


#### Step 3 - Summarize the data 

```{r}
weekend_act <- tapply(new_act_we$steps, new_act_we$interval, mean)
weekday_act <- tapply(new_act_wd$steps, new_act_wd$interval, mean)

```

#### Step 4 - Plot the Weekend and Weekday data adjacently. 

```{r}
par(mfrow = c(1,2))

plot(y = weekend_act, x = names(weekend_act), type = "l", xlab = "5 minutes interval",
     main = "Daily Activtiy Pattern in Weekend", 
     ylab = "Average number of steps")
plot(y = weekday_act, x = names(weekday_act), type = "l", xlab = "5 minutes interval",
     main = "Daily Activtiy Pattern during Weekdays", 
     ylab = "Average number of steps")
```


